---
title: "Predicting Prices of Used Cars"
author: "Vijay Kumar"
date: "6/3/2021"
output:
  pdf_document: default
  word_document: default
  always_allow_html: yes
  html_document:
    df_print: paged
---

# Introduction

The file ToyotaCorolla.csv contains data on used cars (Toyota Corolla) on sale during late summer of 2004 in the Netherlands. It has 1436 records containing details on 38 attributes, including Price, Age, Kilometers, HP, and other specifications. The goal in this assignment is to predict the price of a used Toyota Corolla based on its specifications. We will split the data into training (50%), validation (30%), and test (20%) datasets. Thereafter, we will Run a multiple linear regression with the outcome variable Price and predictor variables Age_08_04, KM, Fuel_Type, HP, Automatic, Doors, Quarterly_ Tax, Mfr_Guarantee, Guarantee_Period, Airco, Automatic_airco, CD_Player,
Powered_Windows, Sport_Model, and Tow_Bar.

# 1. What appears to be the three / four most important features:


## 1.1 Load the libraries required


```{r Libraries, echo=TRUE}

library(knitr)
library(ggplot2)
library(corrr)
library(corrplot)
library(rpart)
library(rpart.plot)
library(tidyverse)
library(tidyr)
library(caret)
library(PerformanceAnalytics)
library(forecast)
library(gains)
library(boot)
library(RColorBrewer)
library(dplyr)
library(pivottabler)
library(dummies)
library(multcomp)
library(lift)
library(leaps)
library(GGally)
library(dummies)

```


## 1.2 Load Dataset, Check Structure, Check Dimensions and Null Values


```{r Dataset, echo=TRUE}

car_df <- read.csv("C:/ToyotaCorolla.csv"
               )
rmarkdown::paged_table(car_df)

```


```{r Check Structure, echo=TRUE}

str(car_df)

```

```{r Check Dimension, echo=TRUE}

dim(car_df)

```
```{r Check Null Values, echo=TRUE}

is.null(car_df)
str(car_df)

```


```{r Dummy Variables, echo=TRUE}

car_df.df <- dummy.data.frame(car_df[,-c(1,2,5,6,11,15)])

str(car_df.df)

```


## 1.3 Checking for the most important car specifications for predicting price:


```{r Correlation Plot, echo=TRUE}

df2 = subset(car_df.df, select = c(Price,Age_08_04,KM,Fuel_TypeCNG, Fuel_TypeDiesel,Fuel_TypePetrol ,HP,Automatic,Doors,Quarterly_Tax,Mfr_Guarantee,Guarantee_Period,Airco,CD_Player,Powered_Windows, Sport_Model,Tow_Bar  ) )

m <- cor(df2, use="pairwise.complete.obs")

corrplot(m, method = "square")
```

From the correlation matrix plot above, we can see that four most important features to determine car price are AGE_08_04, KM,CD_PLAYER& Airco. AGE_08_04 & KM are negatively correlated with price. Hence an increase in these parameters led to a a drop in price. However, CD_Player and Airco seem to be positively correlated to price.


## 1.4 Split dataset into training (50%), validation(30%) & test(20%):


```{r Split_dataset, echo=TRUE}

set.seed(1)

spec = c(train = .5, test = .2, validate = .3)

g = sample(cut(
  seq(nrow(df2)), 
  nrow(df2)*cumsum(c(0,spec)),
  labels = names(spec)
))

res = split(df2, g)

```


## 1.5 Fit linear model on the train dataset:


```{r Linear Model, echo=TRUE}

car_price_lm <- lm ( Price ~ ., data = res$train)

summary(car_price_lm)

```

# 2. Use Metrics you consider useful to assess the performance of the model in predicting prices.


```{r Model Performance - test, echo=TRUE}

lm.pred.test <- predict(car_price_lm, res$test)

accuracy(lm.pred.test, res$test$Price)

```


```{r Model Performance - validation, echo=TRUE}

lm.pred.valid <- predict(car_price_lm, res$validate)

accuracy(lm.pred.valid, res$validate$Price)

```


```{r Model Performance - train, echo=TRUE}

lm.pred.train <- predict(car_price_lm, res$train)

accuracy(lm.pred.train, res$train$Price)

```

The adjusted R2 of the model is 0.8533. This is generally a good R2 value as acceptable R2 values are between 0.6 ~ 1. The RMSE of the test data is higher than the RMSE of the train data indicating a possible overfit. Model performance can be improved through PCA and variable reduction.







